Linear Reduction: Parsing and Interpreting Code Without Trees

### 2024-12-05 Paper/paper.tex
Updated paper

Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
\begin{enumerate}
    \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the grammar of the code.
    \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
\end{enumerate}
For example, we can parse the following expression:

@@

We propose a novel method for the process of interpretation, which doesn't use parse trees.
But our algorithm is not limited to this: parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to
instead produce parse trees.

The idea of the algorithm is as follows: the input is a stream of tokens, which are indivisible characters in any language.
Given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
For example $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
So for example:
\[ \mbt{1} \mbt{+} \mbt{2} \mbt{+} \mbt{3} \to \mbt{1,+} \mbt2 \mbt+ \mbt3 \to \mbt{3} \mbt+ \mbt3 \to \mbt{3,+} \mbt3 \to \mbt6 \]
But what about \ttt{1+2*3}?
We'd get
\[ \mbt1 \mbt+ \mbt2 \mbt* \mbt3 \to \mbt{1,+} \mbt2 \mbt* \mbt3 \to \mbt3 \mbt* \mbt3 \to \mbt{3,*} \mbt3 \to \mbt9 \]

@@

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we define a so-called \textit{initial $\beta$-reducer}: $\hb\colon \Sigma\times\Sigma \longto \Sigma$.
This is a function which tells us how to combine tokens together.
And we extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.

@@

Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.
Namely the initial beta-reducer is the function:
\begin{enumerate}
    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for an operator symbol $\circ$,
    \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$,
    \item $\hb(\mbt{x,$\circ$},\mbt{y,$\bullet$})=\mbt{x$\circ$y,$\bullet$}$,
    \item and any other inputs can be given an arbitrary output.
\end{enumerate}
So for example we can now handle expressions of the form
\begin{multline*}
    \mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt{2,+}_1 \mbt3_\infty \to \mbt{2,+}_1 \mbt3_\infty\\
    \to \mbt{2,+}_1 \mbt3_0 \to \mbt5_0
\end{multline*}

@@

In order to handle parentheses, all we need to do is to make a simple change to our algorithm.
Instead of a full function which also tells us how to compute the new priority, the initial $\beta$-reducer will be a partial function:
\[ \hb\colon \Sigma\times\Sigma\partial\Sigma\times(\bbZ\times\bbZ\to\bbZ) \]
Now we define the derived $\beta$-reducer as follows; it is a function $\beta\colon\oS^*\partial\oS^*$ where for $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$:
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho)$ is defined, then $\beta(\xi)=\tau_{\rho(p_1,p_2)}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (which is a type, a priority, and a value).
Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.

Our rules for the derived $\beta$-reducer is similar to before: it is a function $\beta\colon\oP^* \partial \oP^*$ where for $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0(v_n)$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined, then
    \[ \beta(\xi)=\tau_{\rho(p_1,p_2)}(f(v_1,v_2))\sigma^3_{p_3}(v_3)\cdots\sigma^n_{p_n}(v_n) \]
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}(v_1)\beta(\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n))$.
\end{enumerate}

@@

@@

\section{Producing Parse Trees}

>>>>

Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
\begin{enumerate}
    \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the code according to the language's grammar.
    \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
\end{enumerate}
For example, we can parse the following expression:

@@

We propose a novel method for the process of interpretation, which doesn't use parse trees.
But our algorithm is not limited to this, and can be altered to produce parse trees as well (see \S\ref{producing-parse-trees}).
This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.

The idea of the algorithm is as follows: the input is a stream of tokens, which are indivisible characters in any language.
Given a string of tokens $\mb{\sigma^1}\cdots\mb{\sigma^n}$ (for readability, tokens are written in boxes), we form rules for combining two tokens.
For example, $\mb{\ttt1}$ and $\mb{\ttt+}$ can be combined together to give a token $\mb{\ttt{1,+}}$.
Then if we follow a token of the form $\mb{\ttt{x,+}}$ with a token of the form $\mb{\ttt y}$, the tokens are combined to form a token $\mb{\ttt{x+y}}$.
So for example:
\[ \mbt{1} \mbt{+} \mbt{2} \mbt{+} \mbt{3} \to \mbt{1,+} \mbt2 \mbt+ \mbt3 \to \mbt{3} \mbt+ \mbt3 \to \mbt{3,+} \mbt3 \to \mbt6 \]
But what about \ttt{1+2*3}?
We'd get
\[ \mbt1 \mbt+ \mbt2 \mbt* \mbt3 \to \mbt{1,+} \mbt2 \mbt* \mbt3 \to \mbt3 \mbt* \mbt3 \to \mbt{3,*} \mbt3 \to \mbt9 \]

@@

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
tells us how to combine tokens together.
We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.

@@

Notice that since in our reduction we get pairs of tokens in the form $\mbt{x,*},\mbt{y,+}$, we need to add extra rules to handle these cases.
Specifically here we have $\hb(\mbt{x,*},\mbt{y,+})=\mbt{x*y,+}$.

So let us define $\Sigma$ to consist of tokens of the form $\mbt{x}$ for numbers $\ttt x$, $\mb\circ$ for operator symbols $\circ$ (addition,
multiplication, etc.), and $\mbt{x,$\circ$}$.
Then we define the initial $\beta$-reducer by
\begin{enumerate}
    \item $\hb(\mbt x,\mb\circ)=\mbt{x,$\circ$}$ for operator symbols $\circ$,
    \item $\hb(\mbt{x,$\circ$},\mbt y)=\mbt{x$\circ$y}$,
    \item $\hb(\mbt{x,$\circ$},\mbt{y,$\bullet$})=\mbt{x$\circ$y,$\bullet$}$,
    \item and any other inputs can be given an arbitrary output.
\end{enumerate}
So for example we can now handle expressions of the form
\begin{multline*}
    \mbt1_\infty \mbt*_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt2_\infty \mbt+_1 \mbt3_\infty \to \mbt{1,*}_2 \mbt{2,+}_1 \mbt3_\infty \to \mbt{2,+}_1 \mbt3_\infty\\
    \to \mbt{2,+}_1 \mbt3_0 \to \mbt5_0
\end{multline*}

@@

In order to handle parentheses, all we need to do is to make a simple change to our algorithm.
Instead of a full function which also tells us how to compute the new priority, the initial $\beta$-reducer will be a partial function:
\[ \hb\colon \Sigma\times\Sigma\partial\Sigma\times(\bbZ\times\bbZ\to\bbZ) \]
We add a function to compute the new priority of the token to the output of the initial $\beta$-reducer instead of taking the priorities as inputs.
This is because the new token computed should not be dependent on the priorities of the input tokens, as this just adds extra unnecessary
complexity.

Now we define the derived $\beta$-reducer as follows; it is a function $\beta\colon\oS^*\partial\oS^*$ where for $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$:
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho)$ is defined, then $\beta(\xi)=\tau_{\rho(p_1,p_2)}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

We define our alphabet $\Sigma$ to be the same as in the previous section.

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
Our alphabet of types is $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$, define $\oP=\oS\times\values$ to be the set of all possible terms (a term consists of a type, a priority, and a value).
Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.

Our rules for the derived $\beta$-reducer are similar to before: it is a function $\beta\colon\oP^* \partial \oP^*$ where for $\xi=\sigma^1_{p_1}(v_1)\cdots\sigma^n_{p_n}(v_n)$
\begin{enumerate}
    \item if $n=1$ then $\beta(\xi)=\sigma^n_0(v_n)$,
    \item if $p_1\geq p_2$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined, then
    \[ \beta(\xi)=\tau_{\rho(p_1,p_2)}(f(v_1,v_2))\sigma^3_{p_3}(v_3)\cdots\sigma^n_{p_n}(v_n) \]
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}(v_1)\beta(\sigma^2_{p_2}(v_2)\cdots\sigma^n_{p_n}(v_n))$.
\end{enumerate}

@@

So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
So it is converted to
\[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
The derivation is then
\begin{gather*}
    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \RPAREN\NUM_0(3)\ \OP_1(+)\ \NUM_\infty(4) \\
    \OP\NUM_2(2,*)\ \LPAREN_\infty\ \RPAREN\NUM_0(4)\ \OP_1(+)\ \NUM_\infty(4) \\
    \OP\NUM_2(2,*)\ \NUM_\infty(4)\ \OP_1(+)\ \NUM_\infty(4) \\
    \OP\NUM_2(2,*)\ \OP\NUM_1(4,+)\ \NUM_\infty(4) \\
    \OP\NUM_2(8,+)\ \NUM_\infty(4) \\
    \OP\NUM_2(8,+)\ \NUM_0(4) \\
    \NUM_0(12)
\end{gather*}

@@

\section{Producing Parse Trees} \label{producing-parse-trees}
